---
title: "Logfile_ivar"
output: html_document
date: "2025-04-23"
---

# Introductie

## Original Research
Fibroblast-derived osteoglycin promotes epithelial cell repair.
van der Koog L, Woest ME, Gorter IC, Verschut V, Elferink RAB, Zuidhof AB, Nugraha DF,
Koloko Ngassie ML, Bos SIT, Dhakad D, Wolters JC, Horvatovich PL, Prakash YS, Timens W,
Yildirim ÖA, Brandsma CA, Frijlink HW, Nagelkerke A, Gosens R. NPJ Regen Med. 2025 Mar
25;10(1):16. doi: 10.1038/s41536-025-00404-3.
PMID: 40133363. https://doi.org/10.1038/s41536-025-00404-3

## library
```{r}
# transcriptomics
library("tidyr")
library("dplyr")
library("ggplot2")
library("BiocManager")
library("DESeq2")
library("apeglm")
library("EnhancedVolcano")
library("Biobase")
library("BiocGenerics")
library("S4Vectors")
library("clusterProfiler")
library("org.Hs.eg.db")
library("org.Mm.eg.db")
library("AnnotationDbi")
library("mdatools")
library("paletteer")
```

### reading
23-4-2025


### Reading in the data

```{r}
transcript_data <- read.csv("../data/GSE271272_rawcounts.csv")
head(transcript_data)
```

```{r}
protomic_data <- readxl::read_excel("../data/2021-36_2021-42_SN16withHELAlib_MS2Quant_Qclassic_LocalNorm_NoImp.xlsx")
head(protomic_data)
```
In the first row you can see that the amount of proteins cought in the mass spec differs per sample widh could impact the outcome becouse the SF group seems to have more 

https://www.youtube.com/watch?v=JPpKL1uzE0I 10 min timestamp

why most publisted researchers are false by john
partialy squired discrimant analyse voor 1 en beide datasets
svd
fisher exact test

Uitleg en doel PLSDA
als doel feature selection?
visualisatie patronen die het meest relevant zijn in een groep 

Uitleg en doel SVD


Cant use gene id's as rownames becouse it contains dubplicates
```{r}
proteomic_EV <- data.frame(as.integer(protomic_data$`2021-36_EV1`), as.integer(protomic_data$`2021-42_EV1`), as.integer(protomic_data$`2021-42_EV2`), as.integer(protomic_data$`2021-42_EV3`), as.integer(protomic_data$`2021-42_EV4`))

colnames(proteomic_EV) <- c("36_EV1","42_EV1","42_EV2","42_EV3","42_EV4")

proteomic_EV$EV_na_count <- apply(is.na(proteomic_EV), 1, sum)

head(proteomic_EV)
```

```{r}
proteomic_SF <- data.frame(as.integer(protomic_data$`2021-36_SF1`), as.integer(protomic_data$`2021-36_SF2`), as.integer(protomic_data$`2021-36_SF3`), as.integer( protomic_data$`2021-42_SF1`),as.integer(protomic_data$`2021-42_SF2`))
group_2 <- c("36_SF1","36_SF2","36_SF3","42_SF1","42_SF2")
colnames(proteomic_SF) <- group_2

proteomic_SF$SF_na_count <- apply(is.na(proteomic_SF), 1, sum)

head(proteomic_SF)
```


```{r}
new_frame <- data.frame(protomic_data$PG.ProteinGroups, protomic_data$PG.Genes, proteomic_EV, proteomic_SF)
#head(new_frame)
#length(unique(test$protomic_data.PG.Genes))
#length(test$protomic_data.PG.Genes)
#dublist <- duplicated(test$protomic_data.PG.Genes)
#which(dublist)
new_frame[923,2] <- "TMPO_ISO"
row.names(new_frame) <- new_frame$protomic_data.PG.Genes
# new_frame[923,]
new_frame <- new_frame[-1,]
prot_frame <- subset(new_frame, SF_na_count < 2)
prot_frame <- subset(prot_frame, EV_na_count < 2)
prot_frame <- prot_frame[,-1]
prot_frame <- prot_frame[,-1]
prot_frame <- prot_frame[,-6]
prot_frame <- prot_frame[,-11]
prot_frame
```

imputatie based on mean

```{r mutation}
pre_imp_ev <- prot_frame[,1:5]
pre_imp_ev$gene <- row.names(prot_frame)
#head(pre_imp_ev)
imp_ev <- pre_imp_ev %>%
  mutate(rowMean = as.integer(base::rowMeans(select(., -gene), na.rm = T)),
         across(starts_with("X"),
                ~ coalesce(., rowMean))) %>%
  select(-rowMean)
#sf
pre_imp_sf <- prot_frame[,6:10]
pre_imp_sf$gene <- rownames(prot_frame)
imp_sf <- pre_imp_sf %>%
  mutate(rowMean = as.integer(rowMeans(select(., -gene), na.rm = T)),
         across(starts_with("X"),
                ~ coalesce(., rowMean))) %>%
  select(-rowMean)
imp_ev <- imp_ev[,-6]
imp_sf <- imp_sf[,-6]
imputed_prot <- data.frame(imp_ev, imp_sf)
#rownames(imputed_prot) <- row.names(prot_frame)
head(imputed_prot)
#imputed_prot
```


meta data
```{r}
meta_data <- data_frame(sample = c("X36_EV1","X42_EV1","X42_EV2","X42_EV3","X42_EV4","X36_SF1","X36_SF2","X36_SF3","X42_SF1","X42_SF2"),
                        group = c("Extracellular vesicles ","Extracellular vesicles ","Extracellular vesicles ","Extracellular vesicles ","Extracellular vesicles ","Soluble factors ","Soluble factors ","Soluble factors ","Soluble factors ","Soluble factors ")
                        )
meta_data
```

```{r}
dds <- DESeqDataSetFromMatrix(countData = imputed_prot,
                              colData = meta_data,
                              design = ~ group)
dds
```

```{r}
deseq <- DESeq(dds)
results(deseq)
```

```{r}
resultaat_deseq <- results(deseq, contrast = c("group", "EV","SF"))
resultaat_deseq
```


```{r}
deseq_shrink <- lfcShrink(deseq, coef = "group_SF_vs_EV", type = "apeglm")
deseq_shrink
```

```{r}
resultaat_p_01<- sum(resultaat_deseq$padj < 0.05, na.rm=TRUE)
resultaat_p_01
```

```{r}
deseq_05 <- results(deseq, alpha=0.05)
summary(deseq_05)
```


```{r}
plotMA(resultaat_deseq, ylim=c(-10,10))
```

```{r}
plotMA(deseq_shrink, ylim=c(-10,10))
```


```{r}
EnhancedVolcano(deseq_shrink, x="log2FoldChange", y = "padj",lab = rownames(deseq_shrink), title = "young versus old", pCutoff = 0.05,FCcutoff = 2, ylim = c(0,50),legendPosition = "none",subtitle = "FDR <= 0.05 and absolute FC >= 2")
```

error
```{r}
vsd <- vst(dds, blind = FALSE, nsub = 841)
head(assay(vsd))
```

# pca

```{r}
plotPCA(vsd, intgroup="group", ntop = 841, pcsToUse = c(1,2), returnData = FALSE)
```



```{r}
plotPCA(vsd, intgroup="group", ntop = 841, pcsToUse = c(1,3), returnData = FALSE)+
  ggtitle("PCA of Proteomics data") +
  paletteer::scale_colour_paletteer_d("ggprism::colorblind_safe") +
  theme_classic() +
  theme(panel.grid.major = element_line(colour = "lightgrey"),
        legend.box.background = element_rect(color="grey", size=0.5),
        legend.box.margin = margin(1, 1, 1, 1),
        scale_fill_discrete(labels = c("Extracellular vesicles","Soluble factors ")),
        plot.title = element_text(hjust = 0.5, face = "bold"))

```

# loadings
```{r}
prot_pca <- plotPCA(vsd, intgroup="group", ntop = 841, pcsToUse = c(1,3), returnData = TRUE)
prot_pca
```

```{r}
#imputed_prot
test <- imputed_prot
#colnames(test) <- "group"
colnames(test) <- c("EV","EV","EV","EV","EV","SF","SF","SF","SF","SF")
#test
test_pca <- prcomp(t(test), center = TRUE, scale. = TRUE)
test_pca$x
```

```{r}
x <- data.frame(test_pca$rotation)
pc1 <- x[order(x$PC1, decreasing = TRUE),1, drop = FALSE]
pc2 <- x[order(x$PC2, decreasing = TRUE),2, drop = FALSE]
pc3 <- x[order(x$PC3, decreasing = TRUE),3, drop = FALSE]
pc2
```


# combine valt weg omdat het niet klopt

```{r}
transcript_control_set <- data.frame(transcript_data$Sample, transcript_data$M1_ctrl_Epcam, transcript_data$M2_ctrl_Epcam, transcript_data$M3_ctrl_Epcam, transcript_data$M4_ctrl_Epcam, transcript_data$M1_ctrl_CCL, transcript_data$M2_ctrl_CCL, transcript_data$M3_ctrl_CCL, transcript_data$M4_ctrl_CCL)
colnames(transcript_control_set) <- c("Gene","M1_EP","M2_EP","M3_EP","M4_EP","M1_CC","M2_CC","M3_CC","M4_CC")
head(transcript_control_set)
```
 ACCNUM, ALIAS, ENSEMBL, ENSEMBLPROT, ENSEMBLTRANS, ENTREZID, ENZYME, EVIDENCE, EVIDENCEALL, GENENAME, GENETYPE, GO, GOALL, IPI, MAP, OMIM, 
  conda install bioconda::bioconductor-paletteer
  enzyme kan 900 maar veel dupli
  gene name kan met 318 meer betrouwbaar
```{r}

impute_p_gene <- imputed_prot
impute_p_gene$Gene <- rownames(imputed_prot)

#ts_pr_dataset <- merge(impute_p_gene, transcript_control_set, by.x = "Gene", by.y = "Gene")

#x <- match(impute_p_gene$Gene, ts_pr_dataset$Gene)
#x

#head(ts_pr_dataset)

prot_ali <- bitr(impute_p_gene$Gene, fromType = "ALIAS",
                   toType = "GENENAME",
                   OrgDb = org.Hs.eg.db)
ts_ali <- bitr(transcript_control_set$Gene, fromType = "ALIAS", toType = "GENENAME", OrgDb = org.Mm.eg.db)
prot_ali
```

```{r}
# merging
nodub_pr_ali <- prot_ali[!duplicated(prot_ali$ALIAS),]
nodub_ts_ali <- ts_ali[!duplicated(ts_ali$ALIAS),]

merged_prot <- merge(impute_p_gene, nodub_pr_ali, by.x = "Gene", by.y = "ALIAS")
merged_ts <- merge(transcript_control_set, nodub_ts_ali, by.x = "Gene", by.y = "ALIAS")
merged_tspr <- merge(merged_prot, merged_ts, by.x = "GENENAME", by.y = "GENENAME")
merged_tspr
```

```{r}
merged_tspr_clean <- merged_tspr[c(1,3,4,5,6,7,8,9,10,11,12,15,16,17,18,19,20,21)]
head(merged_tspr_clean)
```

```{r}
library("ggfortify")
#pca((merged_tspr_clean[,-1]))$loadings
pca <- prcomp(t(merged_tspr_clean[,-1]),scale. = TRUE)
autoplot(pca, label = TRUE)
```

```{r}
pca_loading <- pca$rotation
par(mfrow=c(1,1))
plot(pca_loading[,1:2])
```

PLSDA protomics valt weg
data transponeren
colom word 
```{r}
library("mdatools")
#plsda()
#plda_t <- t(imputed_prot)
vst
```



Tekst 

Data imputation:
During the exploretory data analysis we found a large amount of missing data in the proteomics folderm and, because the proteomics analysis only had 10 samples divided over 2 groups we decided to go forward with the genes that only had a maximum of 1 missing value per group. for the missing value we imputed using the mean value of the group. 

Data analysis and visualisation:	 
All data analysis was preformed using R version 4.5.0., using the tidyverse style guidelines.
For data manipulations we used dplyr version 1.1.4 and tidyr 1.3.1. As for visualsing the data we used ggplot 3.5.2 and paletteer 1.6.0

proteomics pca
Protomics was done in the same way as transcriptomics apart from that the analysis was done with 841 genes and with only 2 groups EV and SF.

pc text
Figure 2 is the PCA plot of the proteomics data based on the 1st and the 3rd prncipal component. It can be observed that most of the variation lies is in the first princepel component. and on the 3rd component the EV group is slightly more spread then the SF group. The reason pc 1 and 3 are visualised is because is shows a more interesting visualisation then pc 1 and 2 with only 3% variance difference in pc 2 and 3.


Discussions
pca loadings
In Table 3 we can see the loadings of the proteomics PCA in decending order of the first 3 components. 2 of the genes we found interesting are:
ACTN4 
ACTN4 is a gene that belongs to the alpha actinins gene famaly that can have mutible functions depending on the cell type.
FBLN2
FBLN2 is a gene that is involved in the develepment of the extracalaire matrix that is present in the reparation of dameged cells
and both have strucural implications that could be interesting for the field of regenerative medecine 

The proteomics PCA (figgure 2) shows that the EV and SF groups are very destinct groups where the first princepel component is the main divider containing 67% of the varaince. The 3rd component however shows a bit more simalarity in veraity among the samples and groups with only the EV group being a bit more spread. In the second principel component had more spread in the EV group and less in the SF group. Among the top 5 genes in the loadings of the first 3 principel componets we found 2 genes that could be intresting for future research.


Points of critiek 
We found that this data was not multi omic compatibel becouse when we went threw the data we discoverd that the proteomic and transcriptomic data have no relation that we could statistcly corralate to one another. this is because they are from diffrent sample types transcriptomics data was from mouse based genes and proteomics from human/mouse based genes. Also transcriptomics was done after the proteomics was alreddy done wich coused the issue that you cannot see effects of the EV and SF treatments from transcriptomics to proteomics.

Lack of data:
Their was an overal lack of data that made us unable to make decisives results and in specific the proteomics data set had allot of missing values wich coused us to throw out allot of the data that was gatherd. because of all of the missing values we chose to impute the data where it was possible but becouse of a lack of supporting data this was limmeting and makes the imputation less relaible.

Covid contamination?

Notes 4-6-2025
uitleg ev sf eigen alinia
plsda componenten uit hoogdimensionale data

verbeteringen text

Multi-omics compatibility 

using
When exploring this data we found out that the datasets where from different sample types. The transcriptomics data was taken from mouse-based genes and the proteomics derived human/mouse-based genes. Aswel as the fact that transcriptomics was performed after proteomics, this results in an inability to see the effects of the EV and SF treatments on the same samples in the transcriptomics and proteomics data simultaneously. For these reasons we concluded that the data was not multi-omics compatible, becouse it doesnt have any samples we can statisticlly corralate to one another.
suggested
The data used for this research is not multi-omics compatible. During the data analysis, we discovered that the data have no relation, making us unable to statistically correlate with one another. This is because the data is taken from different sample types the transcriptomics data were taken from  mouse-based genes and the proteomics data were derived from human/mouse based genes. Furthermore the transcriptomics was preformed after the proteomics, this results in an inability to see the effects of the EV and SF treatments on the same samples in the transcriptomics and proteomics data.


PCA
The proteomics PCA (figure 2) shows that the EV and SF groups are very distinct groups where the first principal component is the main divider containing 67% of the variance. The 3rd component however shows a bit more similarity in variety among the samples. Groups containing only EVs are a bit more spread. In the second principal component there is more spread in the EV group and less in the SF group. 
When looking at the top 5 genes in of the first three principal components there were 2 that could be interesting for future research: ACTN4 and FBLN2 (table 3). Both of these genes have structural implications that could be interesting for the field of regenerative medicine. ACTN4 is a gene that belongs to the alpha actinins gene family that can have multiple functions depending on the cell type (PubChem, n.d.). FBLN2 is a gene that is involved in the development of the extracellular matrix that is present in the reparation of damaged cells (FBLN2 Fibulin 2 [Homo Sapiens (Human)] - Gene - NCBI, n.d.). Finding FBLN2 in the samples is to be expected as it is a part of creating the extracellular matrix that is involved in wound repair. (https://pmc.ncbi.nlm.nih.gov/articles/PMC8002899/) What is noteworthy is that acording to the loadings its one of the key factors that separate EV and SF, meaning that EV's carry signifcantly more of this protein then SF's. this is something that could be taken into consideration when designing treatments around this gene. 


In deze Alinea de genen voor het eerst benoemen. Waarom is het logisch/niet logisch dat dit er uit komt? 


Meer waarom toevoegen. 


```{r}
pca_loadings <- plotPCA(vsd, intgroup="group", ntop = 841, pcsToUse = c(1,3), returnData = TRUE)

pca_loadings$group <- as.factor(c("Extracellular vesicles ","Extracellular vesicles ","Extracellular vesicles ","Extracellular vesicles ","Extracellular vesicles ","Soluble factors ","Soluble factors ","Soluble factors ","Soluble factors ","Soluble factors "))

ggplot(pca_loadings, aes(x = PC1, y= PC3)) + 
  geom_point(aes(colour = group), size=4)+
  ggtitle("PCA of Proteomics data") +
  paletteer::scale_colour_paletteer_d("ggprism::colorblind_safe") +
  theme_classic() +
  theme(panel.grid.major = element_line(colour = "lightgrey"),
        legend.box.background = element_rect(color="grey", size=0.5),
        legend.box.margin = margin(1, 1, 1, 1),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

pca 1,2,3

```{r}
pca_pc2 <- plotPCA(vsd, intgroup="group", ntop = 841, pcsToUse = c(1,2), returnData = TRUE)
pca_loadings$PC2 <- pca_pc2$PC2
#pca_loadings
ggplot(pca_loadings, aes(x = PC1, y= PC3)) + 
  geom_point(aes(colour = PC2, shape = group),size=3) +
  scale_color_gradient(low = "blue", high = "red")+
  ggtitle("PCA of Proteomics data") +
  #paletteer::scale_colour_paletteer_d("ggprism::colorblind_safe") +
  theme_classic() +
  theme(panel.grid.major = element_line(colour = "lightgrey"),
        legend.box.background = element_rect(color="grey", size=0.5),
        legend.box.margin = margin(1, 1, 1, 1),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```






