---
title: "Logfile_ivar"
output: html_document
date: "2025-04-23"
---

# Introductie

## Original Research
Fibroblast-derived osteoglycin promotes epithelial cell repair.
van der Koog L, Woest ME, Gorter IC, Verschut V, Elferink RAB, Zuidhof AB, Nugraha DF,
Koloko Ngassie ML, Bos SIT, Dhakad D, Wolters JC, Horvatovich PL, Prakash YS, Timens W,
Yildirim ÖA, Brandsma CA, Frijlink HW, Nagelkerke A, Gosens R. NPJ Regen Med. 2025 Mar
25;10(1):16. doi: 10.1038/s41536-025-00404-3.
PMID: 40133363. https://doi.org/10.1038/s41536-025-00404-3

## library
```{r}
# transcriptomics
library("tidyr")
library("dplyr")
library("ggplot2")
library("BiocManager")
library("DESeq2")
library("apeglm")
library("EnhancedVolcano")
library("Biobase")
library("BiocGenerics")
library("S4Vectors")
library("clusterProfiler")
library("org.Hs.eg.db")
library("org.Mm.eg.db")
library("AnnotationDbi")
library("mdatools")
library("paletteer")
```

### reading
23-4-2025


### Reading in the data

```{r}
transcript_data <- read.csv("../data/GSE271272_rawcounts.csv")
head(transcript_data)
```

```{r}
protomic_data <- readxl::read_excel("../data/2021-36_2021-42_SN16withHELAlib_MS2Quant_Qclassic_LocalNorm_NoImp.xlsx")
head(protomic_data)
```
In the first row you can see that the amount of proteins cought in the mass spec differs per sample widh could impact the outcome becouse the SF group seems to have more 

https://www.youtube.com/watch?v=JPpKL1uzE0I 10 min timestamp

why most publisted researchers are false by john
partialy squired discrimant analyse voor 1 en beide datasets
svd
fisher exact test

Uitleg en doel PLSDA
als doel feature selection?
visualisatie patronen die het meest relevant zijn in een groep 

Uitleg en doel SVD


Cant use gene id's as rownames becouse it contains dubplicates
```{r}
proteomic_EV <- data.frame(as.integer(protomic_data$`2021-36_EV1`), as.integer(protomic_data$`2021-42_EV1`), as.integer(protomic_data$`2021-42_EV2`), as.integer(protomic_data$`2021-42_EV3`), as.integer(protomic_data$`2021-42_EV4`))

colnames(proteomic_EV) <- c("36_EV1","42_EV1","42_EV2","42_EV3","42_EV4")

proteomic_EV$EV_na_count <- apply(is.na(proteomic_EV), 1, sum)

head(proteomic_EV)
```

```{r}
proteomic_SF <- data.frame(as.integer(protomic_data$`2021-36_SF1`), as.integer(protomic_data$`2021-36_SF2`), as.integer(protomic_data$`2021-36_SF3`), as.integer( protomic_data$`2021-42_SF1`),as.integer(protomic_data$`2021-42_SF2`))
group_2 <- c("36_SF1","36_SF2","36_SF3","42_SF1","42_SF2")
colnames(proteomic_SF) <- group_2

proteomic_SF$SF_na_count <- apply(is.na(proteomic_SF), 1, sum)

head(proteomic_SF)
```


```{r}
new_frame <- data.frame(protomic_data$PG.ProteinGroups, protomic_data$PG.Genes, proteomic_EV, proteomic_SF)
#head(new_frame)
#length(unique(test$protomic_data.PG.Genes))
#length(test$protomic_data.PG.Genes)
#dublist <- duplicated(test$protomic_data.PG.Genes)
#which(dublist)
new_frame[923,2] <- "TMPO_ISO"
row.names(new_frame) <- new_frame$protomic_data.PG.Genes
# new_frame[923,]
new_frame <- new_frame[-1,]
prot_frame <- subset(new_frame, SF_na_count < 2)
prot_frame <- subset(prot_frame, EV_na_count < 2)
prot_frame <- prot_frame[,-1]
prot_frame <- prot_frame[,-1]
prot_frame <- prot_frame[,-6]
prot_frame <- prot_frame[,-11]
prot_frame
```

imputatie based on mean

```{r}
pre_imp_ev <- prot_frame[,1:5]
pre_imp_ev$gene <- row.names(prot_frame)
#head(pre_imp_ev)
imp_ev <- pre_imp_ev %>%
  mutate(rowMean = as.integer(rowMeans(select(., -gene), na.rm = T)),
         across(starts_with("X"),
                ~ coalesce(., rowMean))) %>%
  select(-rowMean)

#sf
pre_imp_sf <- prot_frame[,6:10]
pre_imp_sf$gene <- rownames(prot_frame)
imp_sf <- pre_imp_sf %>%
  mutate(rowMean = as.integer(rowMeans(select(., -gene), na.rm = T)),
         across(starts_with("X"),
                ~ coalesce(., rowMean))) %>%
  select(-rowMean)
imp_ev <- imp_ev[,-6]
imp_sf <- imp_sf[,-6]
imputed_prot <- data.frame(imp_ev, imp_sf)
#rownames(imputed_prot) <- row.names(prot_frame)
head(imputed_prot)
#imputed_prot
```


meta data
```{r}
meta_data <- data_frame(sample = c("X36_EV1","X42_EV1","X42_EV2","X42_EV3","X42_EV4","X36_SF1","X36_SF2","X36_SF3","X42_SF1","X42_SF2"),
                        group = c("EV","EV","EV","EV","EV","SF","SF","SF","SF","SF")
                        )
meta_data
```

```{r}
dds <- DESeqDataSetFromMatrix(countData = imputed_prot,
                              colData = meta_data,
                              design = ~ group)
dds
```

```{r}
deseq <- DESeq(dds)
results(deseq)
```

```{r}
resultaat_deseq <- results(deseq, contrast = c("group", "EV","SF"))
resultaat_deseq
```


```{r}
deseq_shrink <- lfcShrink(deseq, coef = "group_SF_vs_EV", type = "apeglm")
deseq_shrink
```

```{r}
resultaat_p_01<- sum(resultaat_deseq$padj < 0.05, na.rm=TRUE)
resultaat_p_01
```

```{r}
deseq_05 <- results(deseq, alpha=0.05)
summary(deseq_05)
```


```{r}
plotMA(resultaat_deseq, ylim=c(-10,10))
```

```{r}
plotMA(deseq_shrink, ylim=c(-10,10))
```


```{r}
EnhancedVolcano(deseq_shrink, x="log2FoldChange", y = "padj",lab = rownames(deseq_shrink), title = "young versus old", pCutoff = 0.05,FCcutoff = 2, ylim = c(0,50),legendPosition = "none",subtitle = "FDR <= 0.05 and absolute FC >= 2")
```

error
```{r}
vsd <- vst(dds, blind = FALSE, nsub = 841)
head(assay(vsd))
```

# pca

```{r}
plotPCA(vsd, intgroup="group", ntop = 841, pcsToUse = c(1,2), returnData = FALSE)
```


```{r}
plotPCA(vsd, intgroup="group", ntop = 841, pcsToUse = c(1,3), returnData = FALSE)+
  ggtitle("PCA of Proteomics data") +
  paletteer::scale_colour_paletteer_d("ggprism::colorblind_safe") +
  theme_classic() +
  theme(panel.grid.major = element_line(colour = "lightgrey"),
        legend.box.background = element_rect(color="grey", size=0.5),
        legend.box.margin = margin(1, 1, 1, 1),
        plot.title = element_text(hjust = 0.5, face = "bold"))

```

# loadings
```{r}
prot_pca <- plotPCA(vsd, intgroup="group", ntop = 841, pcsToUse = c(1,3), returnData = TRUE)
prot_pca
```

```{r}
#imputed_prot
test <- imputed_prot
#colnames(test) <- "group"
colnames(test) <- c("EV","EV","EV","EV","EV","SF","SF","SF","SF","SF")
#test
test_pca <- prcomp(t(test), center = TRUE, scale. = TRUE)
test_pca$x
```

```{r}
x <- data.frame(test_pca$rotation)
pc1 <- x[order(x$PC1, decreasing = TRUE),1, drop = FALSE]
pc2 <- x[order(x$PC2, decreasing = TRUE),2, drop = FALSE]
pc3 <- x[order(x$PC3, decreasing = TRUE),3, drop = FALSE]
pc2
```


# combine valt weg omdat het niet klopt

```{r}
transcript_control_set <- data.frame(transcript_data$Sample, transcript_data$M1_ctrl_Epcam, transcript_data$M2_ctrl_Epcam, transcript_data$M3_ctrl_Epcam, transcript_data$M4_ctrl_Epcam, transcript_data$M1_ctrl_CCL, transcript_data$M2_ctrl_CCL, transcript_data$M3_ctrl_CCL, transcript_data$M4_ctrl_CCL)
colnames(transcript_control_set) <- c("Gene","M1_EP","M2_EP","M3_EP","M4_EP","M1_CC","M2_CC","M3_CC","M4_CC")
head(transcript_control_set)
```
 ACCNUM, ALIAS, ENSEMBL, ENSEMBLPROT, ENSEMBLTRANS, ENTREZID, ENZYME, EVIDENCE, EVIDENCEALL, GENENAME, GENETYPE, GO, GOALL, IPI, MAP, OMIM, 
  conda install bioconda::bioconductor-paletteer
  enzyme kan 900 maar veel dupli
  gene name kan met 318 meer betrouwbaar
```{r}

impute_p_gene <- imputed_prot
impute_p_gene$Gene <- rownames(imputed_prot)

#ts_pr_dataset <- merge(impute_p_gene, transcript_control_set, by.x = "Gene", by.y = "Gene")

#x <- match(impute_p_gene$Gene, ts_pr_dataset$Gene)
#x

#head(ts_pr_dataset)

prot_ali <- bitr(impute_p_gene$Gene, fromType = "ALIAS",
                   toType = "GENENAME",
                   OrgDb = org.Hs.eg.db)
ts_ali <- bitr(transcript_control_set$Gene, fromType = "ALIAS", toType = "GENENAME", OrgDb = org.Mm.eg.db)
prot_ali
```

```{r}
# merging
nodub_pr_ali <- prot_ali[!duplicated(prot_ali$ALIAS),]
nodub_ts_ali <- ts_ali[!duplicated(ts_ali$ALIAS),]

merged_prot <- merge(impute_p_gene, nodub_pr_ali, by.x = "Gene", by.y = "ALIAS")
merged_ts <- merge(transcript_control_set, nodub_ts_ali, by.x = "Gene", by.y = "ALIAS")
merged_tspr <- merge(merged_prot, merged_ts, by.x = "GENENAME", by.y = "GENENAME")
merged_tspr
```

```{r}
merged_tspr_clean <- merged_tspr[c(1,3,4,5,6,7,8,9,10,11,12,15,16,17,18,19,20,21)]
head(merged_tspr_clean)
```

```{r}
library("ggfortify")
#pca((merged_tspr_clean[,-1]))$loadings
pca <- prcomp(t(merged_tspr_clean[,-1]),scale. = TRUE)
autoplot(pca, label = TRUE)
```

```{r}
pca_loading <- pca$rotation
par(mfrow=c(1,1))
plot(pca_loading[,1:2])
```

PLSDA protomics valt weg
data transponeren
colom word 
```{r}
library("mdatools")
#plsda()
#plda_t <- t(imputed_prot)
```












