---
title: "Logfile_Kasthury"
author: "Kasthury Inparajah"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: yes
lang: "en"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

# Proteomics
## Introduction of this project
What are we going to do in this project?

## The article
Write about the article, some background information, keynotes

## Data
What is the data and where is it from?

## 23-4-2025 
### Finding related articles
write about assignments and found articles and overview of it

## 24-4-2025
**"High-throughput proteomics: a methodological mini-review"**
**"Understudied proteins: opportunities and challenges for functional proteomics"**

## Install libraries
```{r Install packages, echo=FALSE}

#install.packages("readxl")
#install.packages("imputeLCMD")
#options(repos = c(CRAN = "https://cloud.r-project.org"))

```

```{r PCAtools download, echo=FALSE}

  if (!requireNamespace('BiocManager', quietly = TRUE))
    install.packages('BiocManager')

  BiocManager::install('PCAtools')
  
```

```{r ClusterProfiler download, echo=FALSE}

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("clusterProfiler")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("org.Mm.eg.db")
BiocManager::install("ReactomePA")

```

```{r Install pathview package}

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("pathview")

```

## Load libraries
```{r Load libraries, echo=FALSE}

library(dplyr)
library(tidyr)
library(readxl)
library(clusterProfiler)
library(org.Hs.eg.db)
library(org.Mm.eg.db)
library(ReactomePA)
library(enrichplot)
library(DESeq2) 

```

## EDA
### **Transcriptomics**
```{r Transcriptomics data, echo=FALSE}

data_transcriptomics <- read.csv("../data/GSE271272_rawcounts.csv")
head(data_transcriptomics)

```

### **Proteomics**
```{r Proteomics data, echo=FALSE}

data_proteomics <- readxl::read_excel("../data/2021-36_2021-42_SN16withHELAlib_MS2Quant_Qclassic_LocalNorm_NoImp.xlsx")
head(data_proteomics)

```
It is noticeable that a lot of values are "Filtered". This makes it difficult to include these data when we analyse the dataset, thus why we need to impute these values.

Besides the filtered data, some samples with numerical values, e.g. from column *2021-36_EV1* are characters instead of numbers.

Now we need to check if the data has any duplicates for the proteins or genes columns. This is crucial because duplicates might influence future analysis and referring to a specific gene becomes almost impossible. 

```{r Check for any duplicates in genes, echo=FALSE}

# Check for duplicates in this columns

# Genes
genes <- data_proteomics$PG.Genes
any(duplicated(genes))

```

The check for duplicates in the genes has returned positive, this means that the gene column has duplicates. 

```{r Check for any duplictas in proteins, echo=FALSE}

# Check for duplicates in this columns

# Proteins
proteins <- data_proteomics$PG.ProteinGroups
any(duplicated(proteins))

```
The check for duplicates in the proteins has returned negative, this means that the proteins column does not contain any duplicates. 

Let's see which gene(s) are duplicated and how many times:

```{r Duplicated genes, echo=FALSE}

# Which gene(s) are duplicated?
duplicates <- genes[duplicated(genes)]
unique(duplicates)

# How many times it is duplicated?
table(genes[genes %in% duplicates])

```

The gene "TMPO" is 2x duplicated. As a solution for this problem, we can make unique rownames with the gene name added by the times it is duplicated.

```{r Unique rownames, echo=FALSE}

# Change to df
data_proteomics <- as.data.frame(data_proteomics)

# Unique genename
unique_genes <- make.unique(data_proteomics$PG.Genes)
rownames(data_proteomics) <- unique_genes

```

Let's make a subset of the dataset divided in EV (extracellular vesicles) and SF (soluble factors).

```{r Subset EV & SF, echo=FALSE}

# EV
ev_cols <- colnames(data_proteomics)[grepl("PG.ProteinGroups|PG.Genes|_EV", colnames(data_proteomics))]
ev_data <- data_proteomics[, ev_cols]

# SF
sf_cols <- colnames(data_proteomics)[grepl("PG.ProteinGroups|PG.Genes|_SF", colnames(data_proteomics))]
sf_data <- data_proteomics[, sf_cols]

```

We now have to remove the metadata from the subset for further analysis e.g. PCA plots. 

```{r Remove metadata, echo=FALSE}

# Remove metadata EV and SF and counts
rm_metadata_ev <- ev_data[, -c(1,2)]
rm_metadata_sf <- sf_data[, -c(1,2)]

```

After we have removed the metadata, we have to change the datatype to numeric. But we first need to change the filtered values to NA.

```{r Change "Filtered" to "NA", echo=FALSE}

# For EV
rm_metadata_ev[rm_metadata_ev == "Filtered"] <- NA
rm_metadata_ev[rm_metadata_ev == ""] <- NA

# For SF
rm_metadata_sf[rm_metadata_sf == "Filtered"] <- NA
rm_metadata_sf[rm_metadata_sf == ""] <- NA

```

Since the dataset has characters instead of numbers, we have to change the datatype.

```{r Change datatype, echo=FALSE}

# For EV and SF
ev_numeric <- as.data.frame(lapply(rm_metadata_ev, as.numeric))
sf_numeric <- as.data.frame(lapply(rm_metadata_sf, as.numeric))

```

Now that we have changed the datatype, we can impute the NA's.

```{r Impute data, echo=FALSE}

# For EV
ev_imputed <- as.data.frame(lapply(ev_numeric, function(x) {
  ifelse(is.na(x), mean(x, na.rm=TRUE), x)
}))

# For SF
sf_imputed <- as.data.frame(lapply(sf_numeric, function(x) {
  ifelse(is.na(x), mean(x, na.rm=TRUE), x)
}))

```

```{r Adding metadata, echo=FALSE}

rownames(ev_imputed) <- unique_genes
rownames(sf_imputed) <- unique_genes

```

We are now going to do a Log2-transformation because the data is usually very skewed (large spread and not normally distributed). Log2 helps normalize the data and makes PCA more reliable.

```{r Log2, echo=FALSE}

pseudo_count <- 1
ev_log2 <- log2(ev_imputed + pseudo_count)
sf_log2 <- log2(sf_imputed + pseudo_count)

```

```{r Filter on the same genes, echo=FALSE}

common_genes <- intersect(rownames(ev_log2), rownames(sf_log2))
ev_log2_common <- ev_log2[common_genes, ]
sf_log2_common <- sf_log2[common_genes, ]

```

```{r Combine subsets, echo=FALSE}

combined_log2 <- cbind(ev_log2_common, sf_log2_common)

```

```{r Transpose, echo=FALSE}

pca_result <- prcomp(t(combined_log2), scale. = TRUE)

```

```{r Plot oud, echo=FALSE}

# Df for plot
pca_df <- data.frame(pca_result$x)
pca_df$Sample <- rownames(pca_df)

# Label sample type
pca_df$Group <- ifelse(grepl("_EV", pca_df$Sample), "EV", "SF")

# Plot
library(ggplot2)
ggplot(pca_df, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "PCA EV & SF samples global")

```

This PCA plot only depicts the difference between EV and SF but not the difference within the groups itself.

```{r Plot, echo=FALSE}

library(ggplot2)

# Df with PCA score and annotation
pca_df <- data.frame(pca_result$x)
pca_df$Sample <- rownames(pca_df)

# Add column with group
pca_df$Group <- ifelse(grepl("_EV", pca_df$Sample), "EV", "SF")
pca_df$Subject <- ifelse(grepl("36", pca_df$Sample), "36", "42")

# Plot
ggplot(pca_df, aes(x=PC1, y=PC2, color=Group, shape=Subject)) +
  geom_point(size=4) +
  theme_minimal() +
  labs(title="PCA EV & SF samples",
       x=paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100,1), "%)"),
       y=paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100,1), "%)"))

```

This PCA plot depicts the PCA scores of the data. Although there is nothing standing out from this plot, the difference between 36_EV and 42_EV is quite big. 

## **Pathway analysis**

Now that we have started with the pathway analysis of the proteins, the first step we have to make is to transpose the names of the genes to the standard ID.

```{r List of genes, echo=FALSE}

# Get all genes
all_genes <- data_proteomics$PG.Genes

# Split genes which are in same row
genes_vector <- unlist(strsplit(all_genes, split = ";"))

# Get the unique genes
unique_genes <- unique(genes_vector)

```

Since our dataset contains data from humans and mice, we are going to do a pathway analysis on both of them. We start with the human pathway. 

```{r Translate gene to Entrez ID for humans, echo=FALSE}

gene_symbols <- unique_genes

# Translate to Entrez ID
entrez_ids_humans <- bitr(gene_symbols, fromType = "SYMBOL",
                   toType = "ENTREZID",
                   OrgDb = org.Hs.eg.db)

# Look at the translation
head(entrez_ids_humans)

```

```{r Vector with ENTREZ ID for humans, echo=FALSE}

# Make vector
entrez_ids_humans <- entrez_ids_humans$ENTREZID
# Make datatype numerical
entrez_ids_humans <- na.omit(as.numeric(entrez_ids_humans))

```

```{r KEGG Pathway analysis for humans, echo=FALSE}

# KEGG pathway analysis for humans
kegg_results_humans <- enrichKEGG(gene = entrez_ids_humans,
                           organism = "hsa",
                           pvalueCutoff = 0.05)

# Results
head(kegg_results_humans)

# Visualisation
barplot(kegg_results_humans, showCategory=15)
dotplot(kegg_results_humans, showCategory=15)

```

```{r Translate gene to Entrez ID for mice, echo=FALSE}

# Translate to Entrez ID
entrez_ids_mice <- bitr(gene_symbols, fromType = "SYMBOL",
                   toType = "ENTREZID",
                   OrgDb = org.Mm.eg.db)

# Look at the translation
head(entrez_ids_mice)

```

```{r Vector with ENTREZ ID for mice, echo=FALSE}

# Make vector
entrez_ids_mice <- entrez_ids_mice$ENTREZID
# Make datatype numerical
entrez_ids_mice <- na.omit(as.numeric(entrez_ids_mice))

```

```{r KEGG Pathway analysis for mice, echo=FALSE}

# KEGG pathway analysis for mice
kegg_results_mice <- enrichKEGG(gene = entrez_ids_mice,
                           organism = "mmu",
                           pvalueCutoff = 0.05)

# Results
head(kegg_results_mice)

# Visualisation
barplot(kegg_results_mice, showCategory=10)
dotplot(kegg_results_mice, showCategory=10)

```

## **20-5-2025**
## **Change of approach analysis**
After discussing with one of our teachers (T. Wassenaar) we have decided to throw the genes away which have more than 1 missing value. The reason why we are doing this, is because we can't make a validate imputation on the missing genes when we miss so much data. Since we only have 5 different values for each gene, it becomes unreasonable for us to use genes of which more than 2 values are missing. Risk with imputating values is that when your expression level are both high and low, only the average of these values will be given for the missing values. This does not represent the data very well, hence why we have changed our approach to this proteomics dataset. 

Our approach after our discussion to analyse the proteomics dataset:
First of all, we are going to make a dataframe, of which we later make the PCA plot of the genes, which have no missing values (this are 1262 genes). Thereafter we will filter genes which have no more than 1 missing value (this are 1575 genes) and make a new dataframe for it. The genes that miss more than 1 value will be thrown away.
The missing values of this dataframe will be imputed with regression. By doing so, the imputed values will have a better representation of the dataset. 

In the end, we will compare both plots to see if there is a difference. 

**NOTE: This above mentioned apprroach is for both EV and SF**

## Dataframe complete genes
As early mentioned before, we are going to make a dataframe of all the genes which have no missing values. 

But first of all, we are going to remove unesseccary data. 

```{r Remove counts, protein and gene row and column, echo=TRUE}

## EV df
## Remove counts (row)
ev_data_rm_counts <- ev_data[-c(1), ]

## Remove protein and genes (column)
ev_data_rm_col_row <- ev_data_rm_counts[-c(1, 2) ]

## SF df
## Remove counts (row)
sf_data_rm_counts <- sf_data[-c(1), ]

## Remove protein and genes (column)
sf_data_rm_col_row <- sf_data_rm_counts[-c(1, 2) ]

```

As we have removed the the uneccessary row and colums, we now have to change the "filtered" or "NaN" values to NA.

```{r Change not numeric values to NA, echo=TRUE}

# For EV
ev_data_rm_col_row[ev_data_rm_col_row == "Filtered"] <- NA
ev_data_rm_col_row[ev_data_rm_col_row == ""] <- NA

# For SF
sf_data_rm_col_row[sf_data_rm_col_row == "Filtered"] <- NA
sf_data_rm_col_row[sf_data_rm_col_row == ""] <- NA

```

Let's extract the gene rows from the df (ev_data_rm_counts and sf_data_rm_counts) to add them later.

```{r Extract gene row from df, eval=TRUE}

sf_genes <- sf_data_rm_counts[c(2)]

ev_genes <- ev_data_rm_counts[c(2)]

```

Now we make an index to see which rows (which are complete) are needed for our new df.

```{r Make index to see which rows are needed with complete data, echo=TRUE}

# Make df numeric
ev_data_num <- as.data.frame(lapply(ev_data_rm_col_row, as.numeric))
sf_data_num <- as.data.frame(lapply(sf_data_rm_col_row, as.numeric))

ev_data_num <- cbind(ev_genes, ev_data_num)
sf_data_num <- cbind(sf_genes, sf_data_num)

# Make an index to see which rows we need
complete_genes_ev <- which(rowSums(is.na(ev_data_num)) == 0)
complete_genes_sf <- which(rowSums(is.na(sf_data_num)) == 0)

# Select the rows and make new df
df_complete_genes_ev <- ev_data_num[complete_genes_ev, , drop = FALSE]
df_complete_genes_sf <- sf_data_num[complete_genes_sf, , drop = FALSE]

```

Let's do the same to genes which miss 1 value.

```{r Make index to see which rows are needed with almost complete data, echo=TRUE}

# Make df numeric
ev_data_num <- as.data.frame(lapply(ev_data_rm_col_row, as.numeric))
sf_data_num <- as.data.frame(lapply(sf_data_rm_col_row, as.numeric))

ev_data_num <- cbind(ev_genes, ev_data_num)
sf_data_num <- cbind(sf_genes, sf_data_num)

# Make an index to see which rows we need
almost_complete_genes_ev <- which(rowSums(is.na(ev_data_num)) == 1)
almost_complete_genes_sf <- which(rowSums(is.na(sf_data_num)) == 1)

# Select the rows and make new df
df_almost_complete_genes_ev <- ev_data_num[almost_complete_genes_ev, , drop = FALSE]
df_almost_complete_genes_sf <- sf_data_num[almost_complete_genes_sf, , drop = FALSE]

```

Since df_almost_complete_genes_ev and df_almost_complete_genes_sf do not have all values, we are going to impute them by taking the mean of the row. 

```{r Impute missing data, eval=TRUE}

# For EV
ev_imputed2 <- as.data.frame(lapply(df_almost_complete_genes_ev, function(x) {
  ifelse(is.na(x), mean(x, na.rm=TRUE), x)
}))

# For SF
sf_imputed2 <- as.data.frame(lapply(df_almost_complete_genes_sf, function(x) {
  ifelse(is.na(x), mean(x, na.rm=TRUE), x)
}))

```

Now we can start with RLE normalisation, but before that, we need to split the gene column from the data and add then as rownames, therefore we also need to make a matrix from the data

```{r Pre-process RLE normalisation by deleting the gene columns, eval=TRUE}

# Complete data

# EV data
ev_complete_matrix <- as.matrix(df_complete_genes_ev[, -1])
rownames(ev_complete_matrix) <- df_complete_genes_ev[, "PG.Genes"]

# SF data
sf_complete_matrix <- as.matrix(df_complete_genes_sf[, -1])
rownames(sf_complete_matrix) <- df_complete_genes_sf[, "PG.Genes"] 

# Imputed data

# EV data
ev_matrix <- as.matrix(ev_imputed2[, -1])
rownames(ev_matrix) <- ev_imputed2[, "PG.Genes"] 

# SF data
sf_matrix <- as.matrix(sf_imputed2[, -1])
rownames(sf_matrix) <- sf_imputed2[, "PG.Genes"] 

```

We can now start with RLE normalisation

```{r Calculate the geometrical mean for the complete data, echo=TRUE}

# Geometrical mean
geo_mean <- function(x) exp(mean(log(x[x > 0])))
ev_geom_means <- apply(ev_complete_matrix, 1, geo_mean)

geo_mean <- function(x) exp(mean(log(x[x > 0])))
sf_geom_means <- apply(sf_complete_matrix, 1, geo_mean)

# Logratio
ev_complete_log_ratios <- log(ev_complete_matrix / ev_geom_means)
sf_complete_log_ratios <- log(sf_complete_matrix / sf_geom_means)

# Size factors
sf_complete_size_factors <- apply(sf_complete_log_ratios, 2, median, na.rm = TRUE)
sf_complete_size_factors <- exp(sf_complete_size_factors)

ev_complete_size_factors <- apply(ev_complete_log_ratios, 2, median, na.rm = TRUE)
ev_complete_size_factors <- exp(ev_complete_size_factors)

# Normalize data
ev_complete_normalized <- sweep(ev_complete_matrix, 2, ev_complete_size_factors, "/")
sf_complete_normalized <- sweep(sf_complete_matrix, 2, sf_complete_size_factors, "/")

# Add genename back to data
ev_complete_normalized_df <- cbind(PG.Genes = rownames(ev_complete_matrix), as.data.frame(ev_complete_normalized))

sf_complete_normalized_df <- cbind(PG.Genes = rownames(sf_complete_matrix), as.data.frame(sf_complete_normalized))

```

```{r Calculate the geometrical mean for the data with imputed values, echo=TRUE}

# Geometrical mean
geo_mean <- function(x) exp(mean(log(x[x > 0])))
ev_geom_means <- apply(ev_matrix, 1, geo_mean)

geo_mean <- function(x) exp(mean(log(x[x > 0])))
sf_geom_means <- apply(sf_matrix, 1, geo_mean)

# Logratio
ev_log_ratios <- log(ev_matrix / ev_geom_means)
sf_log_ratios <- log(sf_matrix / sf_geom_means)

# Size factors
sf_size_factors <- apply(sf_log_ratios, 2, median, na.rm = TRUE)
sf_size_factors <- exp(sf_size_factors)

ev_size_factors <- apply(ev_log_ratios, 2, median, na.rm = TRUE)
ev_size_factors <- exp(ev_size_factors)

# Normalize data
ev_normalized <- sweep(ev_matrix, 2, ev_size_factors, "/")
sf_normalized <- sweep(sf_matrix, 2, sf_size_factors, "/")

# Add genename back to data
ev_normalized_df <- cbind(PG.Genes = rownames(ev_matrix), as.data.frame(ev_normalized))
sf_normalized_df <- cbind(PG.Genes = rownames(sf_matrix), as.data.frame(sf_normalized))

```

Now that we have the data normalised with RLE, we can make a PCA plot. 
Let's start with the data which was already complete and was RLE normalised.

Since EV contains less data compared to SF, we are only going to do a PCA on the genes which are available in both data.

```{r Pre-process complete data for PCA, eval=TRUE}

# Find common genes
common_genes_complete <- intersect(rownames(ev_complete_normalized_df), rownames(sf_complete_normalized_df))

# Only genes available in both df's are conserved
ev_conserved_genes_complete <- ev_complete_normalized_df[common_genes_complete, -1]
sf_conserved_genes_complete <- sf_complete_normalized_df[common_genes_complete, -1]

# Transpose samples to rows
ev_pca_matrix_complete <- t(ev_conserved_genes_complete)
sf_pca_matrix_complete <- t(sf_conserved_genes_complete)

# Add label
ev_group_complete <- data.frame(sample = rownames(ev_pca_matrix_complete), group = "EV")
sf_group_complete <- data.frame(sample = rownames(sf_pca_matrix_complete), group = "SF")

# Combine data
combined_matrix_complete <- rbind(ev_pca_matrix_complete, sf_pca_matrix_complete)
group_info_complete <- rbind(ev_group_complete, sf_group_complete)

# Execute PCA
pca_result_complete <- prcomp(combined_matrix_complete, scale. = TRUE)

# Make PCA df
pca_df_complete <- as.data.frame(pca_result_complete$x)
pca_df_complete$group <- group_info_complete$group
pca_df_complete$sample <- rownames(pca_df_complete)
pca_df_complete$subject <- ifelse(grepl("36", pca_df_complete$sample), "36", "42")

```

Let's make the PCA plot. 

```{r PCA plot complete data, eval=TRUE}

library(ggplot2)

ggplot(pca_df_complete, aes(x = PC1, y = PC2, color = group, label = sample, shape=subject)) +
  geom_point(size = 4) +
  geom_text(vjust = -1.2, size = 3) +
  theme_minimal() +
  labs(title = "PCA EV vs SF (complete data)",
       x = paste0("PC1 (", round(summary(pca_result_complete)$importance[2, 1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_result_complete)$importance[2, 2] * 100, 1), "%)"))

```

Now that we have made a PCA for the complete data, let's make one for the imputed data. 

```{r {r Pre-process imputed data for PCA, eval=TRUE}

# Find common genes
common_genes_imputed <- intersect(rownames(ev_normalized_df), rownames(sf_normalized_df))

# Only genes available in both df's are conserved
ev_conserved_genes_imputed <- ev_normalized_df[common_genes_imputed, -1]
sf_conserved_genes_imputed <- sf_normalized_df[common_genes_imputed, -1]

# Transpose samples to rows
ev_pca_matrix_imputed <- t(ev_conserved_genes_imputed)
sf_pca_matrix_imputed <- t(sf_conserved_genes_imputed)

# Add label
ev_group_imputed <- data.frame(sample = rownames(ev_pca_matrix_imputed), group = "EV")
sf_group_imputed <- data.frame(sample = rownames(sf_pca_matrix_imputed), group = "SF")

# Combine data
combined_matrix_imputed <- rbind(ev_pca_matrix_imputed, sf_pca_matrix_imputed)
group_info_imputed <- rbind(ev_group_imputed, sf_group_imputed)

# Execute PCA
pca_result_imputed <- prcomp(combined_matrix_imputed, scale. = TRUE)

# Make PCA df
pca_df_imputed <- as.data.frame(pca_result_imputed$x)
pca_df_imputed$group <- group_info_imputed$group
pca_df_imputed$sample <- rownames(pca_df_imputed)
pca_df_imputed$subject <- ifelse(grepl("36", pca_df_imputed$sample), "36", "42")

```

Now we are going to make a plot for the imputed data.

```{r PCA plot imputed data, eval=TRUE}

library(ggplot2)

ggplot(pca_df_imputed, aes(x = PC1, y = PC2, color = group, label = sample, shape=subject)) +
  geom_point(size = 4) +
  geom_text(vjust = -1.2, size = 3) +
  theme_minimal() +
  labs(title = "PCA EV vs SF (imputed data)",
       x = paste0("PC1 (", round(summary(pca_result_imputed)$importance[2, 1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(summary(pca_result_imputed)$importance[2, 2] * 100, 1), "%)"))

```

## **Pathway analysis Filtered data**
Just like we made a pathway analysis for the unfiltered data, we are now going to make one for the complete and imputed data. We are going to start with the complete data.

First of all, we have to transpose the names of the genes to the standard ID.

```{r List of genes (complete data), echo=TRUE}

# Get all genes
all_genes_complete <- ev_complete_normalized_df$PG.Genes

# Split genes which are in same row
genes_vector_complete <- unlist(strsplit(all_genes_complete, split = ";"))

# Get the unique genes
unique_genes_complete <- unique(genes_vector_complete)

```

As I have mentioned before, we have mice and human data. We start with the human pathway. 
```{r Translate gene to Entrez ID for humans (complete data), echo=TRUE}

gene_symbols_complete <- unique_genes_complete

# Translate to Entrez ID
entrez_ids_humans_complete <- bitr(gene_symbols_complete, fromType = "SYMBOL",
                   toType = "ENTREZID",
                   OrgDb = org.Hs.eg.db)

# Look at the translation
head(entrez_ids_humans_complete)

```

```{r Vector with ENTREZ ID for humans (complete data), echo=TRUE}

# Make vector
entrez_ids_humans_complete <- entrez_ids_humans_complete$ENTREZID

# Make datatype numerical
entrez_ids_humans_complete <- na.omit(as.numeric(entrez_ids_humans_complete))

```

```{r KEGG Pathway analysis for humans  (complete data), echo=TRUE}

# KEGG pathway analysis for humans
kegg_results_humans_complete <- enrichKEGG(gene = entrez_ids_humans_complete,
                           organism = "hsa",
                           pvalueCutoff = 0.05)

# Results
head(kegg_results_humans_complete)

# Visualisation
barplot(kegg_results_humans_complete, showCategory=15)
dotplot(kegg_results_humans_complete, showCategory=15)

```

And now for the mice.

```{r Translate gene to Entrez ID for mice (complete data), echo=TRUE}

gene_symbols_complete <- unique_genes_complete

# Translate to Entrez ID
entrez_ids_mice_complete <- bitr(gene_symbols_complete, fromType = "SYMBOL",
                   toType = "ENTREZID",
                   OrgDb = org.Mm.eg.db)

# Look at the translation
head(entrez_ids_mice_complete)

```

```{r Vector with ENTREZ ID for mice (complete data), echo=TRUE}

# Make vector
entrez_ids_mice_complete <- entrez_ids_mice_complete$ENTREZID

# Make datatype numerical
entrez_ids_mice_complete <- na.omit(as.numeric(entrez_ids_mice_complete))

```

```{r KEGG Pathway analysis for mice  (complete data), echo=TRUE}

# KEGG pathway analysis for mice
kegg_results_mice_complete <- enrichKEGG(gene = entrez_ids_mice_complete,
                           organism = "mmu",
                           pvalueCutoff = 0.05)

# Results
head(kegg_results_mice_complete)

# Visualisation
barplot(kegg_results_mice_complete, showCategory=15)
dotplot(kegg_results_mice_complete, showCategory=15)

```

Let's do the pathway analysis for the imputed data.

```{r List of genes (imputed data), echo=TRUE}

# Get all genes
all_genes_imputed <- ev_normalized_df$PG.Genes

# Split genes which are in same row
genes_vector_imputed <- unlist(strsplit(all_genes_imputed, split = ";"))

# Get the unique genes
unique_genes_imputed <- unique(genes_vector_imputed)

```

```{r Translate gene to Entrez ID for humans (imputed data), echo=TRUE}

gene_symbols_imputed <- unique_genes_imputed

# Translate to Entrez ID
entrez_ids_humans_imputed <- bitr(gene_symbols_imputed, fromType = "SYMBOL",
                   toType = "ENTREZID",
                   OrgDb = org.Hs.eg.db)

# Look at the translation
head(entrez_ids_humans_imputed)

```

```{r Vector with ENTREZ ID for humans (imputed data), echo=TRUE}

# Make vector
entrez_ids_humans_imputed <- entrez_ids_humans_imputed$ENTREZID

# Make datatype numerical
entrez_ids_humans_imputed <- na.omit(as.numeric(entrez_ids_humans_imputed))

```

```{r KEGG Pathway analysis for humans  (imputed data), echo=TRUE}

# KEGG pathway analysis for humans
kegg_results_humans_imputed <- enrichKEGG(gene = entrez_ids_humans_imputed,
                           organism = "hsa",
                           pvalueCutoff = 0.05)

# Results
head(kegg_results_humans_imputed)

# Visualisation
barplot(kegg_results_humans_imputed, showCategory=10)
dotplot(kegg_results_humans_imputed, showCategory=10)

```

And now for the mice.

```{r Translate gene to Entrez ID for mice (imputed data), echo=TRUE}

gene_symbols_imputed <- unique_genes_imputed

# Translate to Entrez ID
entrez_ids_mice_imputed <- bitr(gene_symbols_imputed, fromType = "SYMBOL",
                   toType = "ENTREZID",
                   OrgDb = org.Mm.eg.db)

# Look at the translation
head(entrez_ids_mice_imputed)

```

```{r Vector with ENTREZ ID for mice (imputed data), echo=TRUE}

# Make vector
entrez_ids_mice_imputed <- entrez_ids_mice_imputed$ENTREZID

# Make datatype numerical
entrez_ids_mice_imputed <- na.omit(as.numeric(entrez_ids_mice_imputed))

```

```{r KEGG Pathway analysis for mice  (imputed data), echo=TRUE}

# KEGG pathway analysis for mice
kegg_results_mice_imputed <- enrichKEGG(gene = entrez_ids_mice_imputed,
                           organism = "mmu",
                           pvalueCutoff = 0.05)

# Results
head(kegg_results_mice_imputed)

# Visualisation
barplot(kegg_results_mice_imputed, showCategory=10)
dotplot(kegg_results_mice_imputed, showCategory=10)

```

We are now going to make a visualisation of the pathway. This code was written by my peer J. Jonker. 

Visualisation pathway human complete data:

```{r Visualisation pathway COVID-19 complete data human, eval=TRUE}
library(pathview)

setwd("visualisation_pathway_proteomics/imputed_data/human")

pathview(
  gene.data = kegg_results_humans_imputed,
  pathway.id = "hsa05171",
  species = "hsa",
  out.suffix = "Coronavirus_disease_COVID-19"
)

# Chemokine signaling pathway
pathview(
  gene.data = kegg_results_humans_imputed,
  pathway.id = "hsa05171",
  species = "hsa",
  out.suffix = "Chemokine"
)

```

Visualisation pathway mice complete data:
```{r Visualisation pathway COVID-19 complete data mice, eval=TRUE}

setwd("visualisation_pathway_proteomics/complete_data/mice")

pathview(
  gene.data = kegg_results_mice_complete,
  pathway.id = "mmu05171",
  species = "mmu",
  out.suffix = "Coronavirus_disease_COVID-19"
)

# Chemokine signaling pathway
pathview(
  gene.data = kegg_results_mice_complete,
  pathway.id = "mmu05171",
  species = "mmu",
  out.suffix = "Chemokine"
)

```

Visualisation pathway mice imputed data:
```{r Visualisation pathway COVID-19 imputed data mice, eval=TRUE}

setwd("visualisation_pathway_proteomics/imputed_data/mice")

pathview(
  gene.data = kegg_results_mice_imputed,
  pathway.id = "mmu05171",
  species = "mmu",
  out.suffix = "Coronavirus_disease_COVID-19"
)

# Chemokine signaling pathway
pathview(
  gene.data = kegg_results_mice_imputed,
  pathway.id = "mmu05171",
  species = "mmu",
  out.suffix = "Chemokine"
)

```
